---
title: "Flight Delay Prediction with ridgereg"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Flight Delay Prediction with ridgereg}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction

We will use the function `ridgereg` to predict flight delays using the `nycflights13` dataset. We will explore the weather data and flight data, create interaction effects, and evaluate the model performance.

```{r setup, include=F}
# Load necessary libraries
library(LinearRegression)
library(nycflights13)
library(dplyr)
library(caret)
```

## Data Preparation

We will remove variables that are not believed to have predictive value.

```{r}
# Load the flights and weather datasets
data("flights")
data("weather")

# Merge flights and weather data on common keys
flights_weather <- flights %>%
  left_join(weather, by = c("year", "month", "day", "hour", "origin"))

# Selecting columns that could have predictive value
flights_weather <- flights_weather %>%
  dplyr::select(
    dep_delay, arr_delay, air_time, distance,
    temp, dewp, humid, wind_speed, wind_gust, precip, visib
  )

# Remove rows with NA values
flights_weather <- na.omit(flights_weather)
```

## Create Interaction Effects

We will create interaction effects that could be significant in predicting delays.

```{r}
# Adding interaction terms
flights_weather <- flights_weather %>%
  mutate(
    wind_speed_precip = wind_speed * precip,
    wind_speed_visib = wind_speed * visib
  )
```

## Split Data into Train, Validation, and Test Sets

We split the data into training (80%), validation (15%), and test (5%) sets.

```{r}
# Setting the seed for reproducibility
set.seed(123)

# Partition the data
trainIndex <- createDataPartition(flights_weather$dep_delay, p = 0.8, list = FALSE)
train_set <- flights_weather[trainIndex, ]
temp_set <- flights_weather[-trainIndex, ]

# Split the remaining data into validation and test sets
valIndex <- createDataPartition(temp_set$dep_delay, p = 0.75, list = FALSE)
validation_set <- temp_set[valIndex, ]
test_set <- temp_set[-valIndex, ]
```

## Train and Evaluate Model

We train ridge regression models for different values of `lambda`, and evaluate their performance on the validation set. We’ll use RMSE as our evaluation metric.

```{r}
# Define a grid of lambda values to test
lambda_grid <- seq(0, 10, by = 0.5)

# Initialize a data frame to store validation RMSE results
rmse_results <- data.frame(lambda = lambda_grid, RMSE = NA)

# Training ridge regression models
for (i in seq_along(lambda_grid)) {
  lambda <- lambda_grid[i]
  # Fit the model on the training data
  ridgereg_model <- ridgereg(dep_delay ~ ., data = train_set, lambda = lambda)

  # Predict on the validation set
  preds <- LinearRegression::predict(ridgereg_model, newdata = validation_set)

  # Calculate RMSE on validation set
  rmse_results$RMSE[i] <- sqrt(mean((validation_set$dep_delay - preds)^2))
}

# Find the optimal lambda value
best_lambda <- rmse_results$lambda[which.min(rmse_results$RMSE)]
```

## Evaluate on Test Set

Using the best lambda value, we train the model on the combined train and validation sets and evaluate it on the test set. The test RMSE with optimal lambda is then printed out.

```{r}
# Combine train and validation sets
train_val_set <- rbind(train_set, validation_set)

# Train final model with best lambda
final_model <- ridgereg(dep_delay ~ ., data = train_val_set, lambda = best_lambda)

# Predict on the test set
test_preds <- LinearRegression::predict(final_model, newdata = test_set)

# Calculate RMSE on the test set
test_rmse <- sqrt(mean((test_set$dep_delay - test_preds)^2))

cat("Test RMSE with Optimal Lambda:", test_rmse, "\n")
```
